Boutiques: a flexible framework for automated application integration in computing platforms

### 2016-11-11 paper.tex
First part of the discussion

@@

\paragraph{Niak.} The Niak pre-processing pipeline, executed with the Pipeline System
for Octave and Matlab (PSOM)~\cite{bellec2012pipeline}, was integrated
in CBRAIN using Boutiques. The integration uses the parallelization
mechanism described in Section~\ref{sec:parallelization} so that even
the invocations processing a single subject can be parallelized. It also
allows CBRAIN to leverage the efficient agent model used in PSOM, as
described in~\cite{GLAT-16}. The integration required some work in
PSOM to facilitate its invocation as a non-interactive command-line
application. The resulting CBRAIN plugin is available at
\url{https://github.com/SIMEXP/cbrain-plugins-psom}. It could in
principle be used in other platforms that support the parallelization
property, although this has not been tested.

@@

\paragraph{SPM.} The Statistical Parameter Mapping toolbox (SPM)~\cite{penny2011statistical}
was exported to Boutiques through a specific \texttt{spm2boutiques}
tool that writes Boutiques manifests from configuration files of SPM
functions (object declarations). \todo{More on that when it is
  completed.}

\paragraph{MICCAI challenges.} Boutiques was also used to port to the VIP platform the 23 pipelines
involved in two computational challenges organized by the MICCAI
conference in 2016, related to the segmentation of multiple-sclerosis
lesions in MR images and of tumor volumes in PET images. The pipeline
developers created Boutiques manifests and wrapped their applications
in Docker containers. The pipelines were ported to VIP and executed on
205 subjects in a few weeks only. Some pipelines had to be adjusted
once ported to the platform for the following reasons:
\begin{enumerate}
\item A pipeline
  required a GPU, which we enabled through the
  \texttt{nvidia-docker}\footnote{\url{https://github.com/NVIDIA/nvidia-docker}}
  tool not supported in Boutiques by default although it could be a possible extension.
\item A pipeline required
  more than 10 GB of data dependencies (atlas data) which exceeded the
  maximum size allowed for Docker containers in our setup; we solved the
  issue by installing the data on a directory of the host server that we
  mounted in the container. This could not possibly be supported by
  Boutiques since it is very dependent on the platform.
\item A pipeline wrote more than 10 GB of intermediate data in a
  temporary directory of the container located on a 2 GB partition. We
  solved the issue by mounting a host directory in the temporary directory.
\end{enumerate}
The pipelines did not use much of the expressiveness features offered
by Boutiques since the instructions given to the challengers required
that all the pipeline parameters were fixed, to facilitate
execution. Two custom properties (\texttt{vip:miccai-challenger-email}
and \texttt{vip:miccai-challenge-team-name}) were added to the
Boutiques manifest to help VIP handle pipeline results for the 2016
MICCAI challenges.

@@

\begin{itemize}
\item Security: containers may be updated after being integrated in the platforms.
\item The manifest specification is getting increasingly complex, but
  this complexity is required to describe real-world
  applications. When something cannot be described using the manifest,
  it is always possible to change the command line format by writing a
  wrapper script for the application. However, when containers are
  used, the wrapper script needs to be added to the container, which
  has some overhead (in particular, the containers provided by the
  developers cannot be used directly).
\item Testing is critical.
\item Workflow engines can be platforms (they can run Boutiques applications), workflows can be applications (they can be described in Boutiques).
\item With Boutiques, developers can port their application once and execute it in several platforms. It removes the technological dependency to a particular platform and facilitates migration. 
\end{itemize}

The \boutiques specification adopts a bottom-up approach where new
features are progressively added when applications or platforms
require them. It is currently supported by \todo{X} platforms and  \todo{Y} applications
were described.

 Although the motivating use cases
were taken from neuroinformatics, our primary field of interest,
nothing prevents the system from being used in a variety of other
fields.

@@

>>>>

\todo{Sorina, Tristan}

@@

\paragraph{Niak.} The Niak pre-processing pipeline, executed with the Pipeline System
for Octave and Matlab (PSOM)~\cite{bellec2012pipeline}, was integrated
in CBRAIN through Boutiques. The integration uses the parallelization
mechanism described in Section~\ref{sec:parallelization} so that even
the invocations processing a single subject can be parallelized. It also
allows CBRAIN to leverage the efficient agent model used in PSOM, as
described in~\cite{GLAT-16}. The integration required some work in
PSOM to facilitate its invocation as a non-interactive command-line
application. The resulting CBRAIN plugin is available at
\url{https://github.com/SIMEXP/cbrain-plugins-psom}. It could in
principle be used in other platforms that support the parallelization
property, although this has not been tested.

@@

\paragraph{SPM.} The Statistical Parameter Mapping toolbox (SPM)~\cite{penny2011statistical}
was exported to Boutiques \ldots \todo{Guillaume}

\paragraph{MICCAI challenges.}
Boutiques was used to integrate 23 pipelines in the VIP platform in
the context of two challenges organized by the MICCAI conference in
2016, related to the segmentation of multiple-sclerosis lesions in MR
images (MSSEG
challenge\footnote{\url{https://portal.fli-iam.irisa.fr/msseg-challenge/overview}})
and of tumor volumes in PET images (PETSEG
challenge\footnote{\url{https://portal.fli-iam.irisa.fr/petseg-challenge/overview}}). The
pipelines were ported to VIP and executed on 205 subjects in a few
weeks only, which would not have been possible without
\boutiques. Some pipelines had to be adjusted manually once ported to
the platform for the following reasons:
\begin{enumerate}
\item A pipeline
  required a GPU, which we enabled through the
  \texttt{nvidia-docker}\footnote{\url{https://github.com/NVIDIA/nvidia-docker}}
  tool not supported in Boutiques by default although it could be a possible extension.
\item A pipeline required more than 10 GB of data dependencies (atlas
  data) which exceeded the maximum size allowed for Docker containers
  in our setup. We solved the issue by installing the data in a
  directory of the host server that we mounted in the container. Such
  platform-specific configurations could not possibly be supported by
  \boutiques.
\item A pipeline wrote more than 10 GB of intermediate data in a
  temporary directory of the container located on a 2 GB partition. We
  solved the issue by mounting a host directory in the temporary directory.
\end{enumerate}
The pipelines did not use much of the expressiveness features offered
by Boutiques since the instructions given to the challengers required
that all the pipeline parameters were hardcoded, to facilitate
execution. Two custom properties (\texttt{vip:miccai-challenger-email}
and \texttt{vip:miccai-challenge-team-name}) were added to the
\boutiques manifest to help post-process results in the specific
context of MICCAI challenges.

@@

With Boutiques, developers can port their application once and execute
it in several platforms. It removes the technological dependency to a
particular platform and facilitates migration.  Although the
motivating use cases were taken from neuroinformatics, our primary
field of interest, nothing prevents the system from being used in a
variety of other fields.

\subsection{\boutiques manifest}

The \boutiques manifest specification allows describing a wide range
of applications, but it is also getting increasingly complex through
additions such as invocation schemas and dependencies among inputs. In
the future, even more features might be added, for instance to allow
for a better description of when optional outputs are produced, i.e.,
which combination of input values is required. Extending the
\boutiques manifest has two main goals. First, accurately describing
application improves validation: incorrect input values and execution
results are more precisely detected when the application manifest is
comprehensive, which is essential to improve the reliability and
user-friendliness of execution platforms. Secondly, a rich manifest
schema reduces the need for custom wrappers to integrate applications
as even convoluted command lines can be described: this is
particularly interesting for applications installed in containers
since adding a wrapper to the execution stack requires an update in
the container image.

Nonetheless, a complex manifest schema has a cost for application
developers and platforms, which we address as follows. For developers,
we maintain the set of mandatory manifest properties as small as
possible so that simple applications can be described in a few lines
only (see Listing~\ref{listing:minimal}). For platforms, we aim at
supporting as many features as possible in the local executor so that
only the following steps need to be implemented to support \boutiques
in a platform, regardless of the complexity of the manifest:
\begin{enumerate}
  \item Input entry: generate the interface to allow users to enter
    input values.
  \item Input validation: create a JSON object from the interface,
    validate it against the invocation schema.
  \item Execution: generate and run the command line from the local
    executor and input JSON object.
  \item Output delivery: from the manifest, identify the output files
    and deliver them to the user.
\end{enumerate}
In particular, advanced validation features such as dependencies
between inputs are embedded in the regular JSON validation performed
at step 2, without requiring the platform to support the related
manifest properties.
\begin{listing}
\begin{minted}[frame=single,
               framesep=3mm,
               linenos=false,
               xleftmargin=21pt,
               tabsize=4]{js}
{
    "name": "echo",
    "tool-version": "1.0",
    "description": "A simple script to test output files",
    "command-line": "echo [PARAM] > output.txt",
    "schema-version": "0.4",
    "inputs": [{
	"id": "param",
	"name": "Parameter",
        "value-key": "[INPUT_FILE]",
	"type": "File"
    }],
    "output-files": [{
	"id": "output_file",
	"name": "Output file",
	"path-template": "output.txt"
    }]

}
\end{minted}
\caption{A minimal \boutiques manifest.} 
\label{listing:minimal}
\end{listing}

\subsection{Workflow support}

The parallelization framework described in~\ref{sec:parallelization}
supports workflows by considering them as both applications and
platforms, which we think is a powerful but underexploited
model. Workflows are applications because they are invoked like any
other script or executable, possibly through a command line. Workflows
are also platforms because they may invoke other applications that
might themselves be described in \boutiques. This model is powerful
because it shields the \boutiques specification from specific workflow
constructs and allows a wide range of workflow engines to be described
uniformly.

\subsection{Limitations}

A few limitations remain that should be addressed in the
future. When

\begin{itemize}
\item Security: containers may be updated after being integrated in the platforms.
\item Testing of applications.
\end{itemize}

@@

The \boutiques specification adopts a bottom-up
approach where new features are progressively added when applications
or platforms require them.
